### 2024-07-02 causal-inference.qmd
20240702 - mediation causal diagrams

Mediation can be divided into two types: [full](#sec-causalDiagramMediationFull) and [partial](#sec-causalDiagramMediationPartial).
In [full mediation](#sec-causalDiagramMediationFull), the mediator(s) fully account for the effect of the predictor variable on the outcome variable.
In [partial mediation](#sec-causalDiagramMediationPartial), the mediator(s) partially but do not fully account for the the effect of the [predictor variable](#sec-correlationalStudy) on the [outcome variable](#sec-correlationalStudy).

@@

@@

>>>>

Mediation can be divided into two types: [full](#sec-causalDiagramMediationFull) and [partial](#sec-causalDiagramMediationPartial).
In [full mediation](#sec-causalDiagramMediationFull), the mediator(s) fully account for the effect of the predictor variable on the outcome variable.
In [partial mediation](#sec-causalDiagramMediationPartial), the mediator(s) partially but do not fully account for the effect of the [predictor variable](#sec-correlationalStudy) on the [outcome variable](#sec-correlationalStudy).

@@

In partial mediation, `X` and `Y` are *not* conditionally independent when accounting for the mediator (`Z`).
In this case, coaching quality (`X`) and fantasy points (`Y`) are still associated when accounting for player preparation (`M`).
In other words, in this example, player preparation is a mechanism that partially but does not fully account for the effect of coaching quality on players' fantasy points.
That is, there are likely other mechanisms, in addition to player preparation, that collectively account for the effect of coaching quality on fantasy points.
For instance, coaching quality could also influence player fantasy points through better play-calling.

As with [full mediation](#sec-causalDiagramMediationPartial), the output indicates that, to obtain an unbiased estimate of the *direct* causal association between coaching quality (`X`) and fantasy points (`Y`) (i.e., the effect that is *not* mediated through intermediate processes), it is necessary to control for player preparation (`M`).

@@

However, as with [full mediation](#sec-causalDiagramMediationPartial), to obtain an unbiased estimate of the *total* causal association between coaching quality (`X`) and fantasy points (`Y`) (i.e., including the portion of the effect that is mediated through intermediate processes), it is important *not* to control for player preparation (`M`).
When the goal is to understand the (total) causal effect of coaching quality (`X`) and fantasy points (`Y`), controlling for a mediator (player preparation; `M`) would be inappropriate because doing so would remove the causal effect, thus artificially reducing the estimate of the association between coaching quality (`X`) and fantasy points (`Y`) [@Lederer2019].
